---
title: "Final Project"
format: html
editor: visual
---

## INF 385T.9 - Data Wrangling Final Project: Does Positive Pop Music Portend Positive Attitudes?

## R Tidyverse

## Load tidyverse

```{r}

library(tidyverse)
```

## Import & Wrangle Dataset 1 - World Happiness Report

References:

<https://dplyr.tidyverse.org/reference/rename.html>

<https://dplyr.tidyverse.org/reference/select.html>

```{r}

## Load World Happiness Report - 2023
WHR23 <- read.csv('/Users/Jaykay/Desktop/INF 385T.9 - Data Wrangling/Final Project Data/WHR23.csv')

WHR23
```

```{r}

## Select Country.name,Ladder.score,Logged.GDP.per.capita, Social.support,Healthy.life.expectancy,
## Freedom.to.make.life.choices,Generosity
Happiness_report_23 <- WHR23 |>
                        select(c(Country.name,
                                 Ladder.score, 
                                 Logged.GDP.per.capita, 
                                 Social.support,
                                 Healthy.life.expectancy,
                                 Freedom.to.make.life.choices,
                                 Generosity))

Happiness_report_23
```

```{r}

## Rename columns for more intuitive understanding
happiness_report_23 <- Happiness_report_23 |>
                        rename(c(country_name = Country.name,
                                 happiness_score = Ladder.score,
                                 gdp_wellbeing = Logged.GDP.per.capita,
                                 social_support = Social.support,
                                 life_exp = Healthy.life.expectancy,
                                 freedom = Freedom.to.make.life.choices,
                                 generosity = Generosity))

happiness_report_23
```

## Import & Wrangle Dataset 2 - Spotify Global Top 50 Song Data (Kaggle)

References:

<https://tidyr.tidyverse.org/reference/separate_wider_delim.html>

<https://dplyr.tidyverse.org/reference/relocate.html>

```{r}

## Load global top spotify song (2023) dataset
universal_top_spotify_songs_2023 <- read.csv('/Users/Jaykay/Desktop/INF 385T.9 - Data Wrangling/Final Project Data/universal_top_spotify_songs_2023.csv')

universal_top_spotify_songs_2023
```

```{r}

## Select name, artists, country, danceability ,energy, loudness, valence, and tempo
spotify_2023 <- universal_top_spotify_songs_2023 |>
                  select(c(name, 
                           artists, 
                           country,
                           danceability,
                           energy, 
                           loudness,
                           valence,
                           tempo))


## Rename name column to title
spotify_2023 <- spotify_2023 |>
                  rename(title = name)

spotify_2023
```

```{r}

## Split the artists column and only keep the first artist
## For smoother joins, change following: 
## Remove any white space before and after the title and artist columns
## Change the case to all lowercase for both title and artist columns
global_spotify_2023 <- spotify_2023 |>
                          mutate(artists = str_to_lower(str_trim(as.character(artists))), 
                                 title = str_to_lower(str_trim(title))) |>
                          mutate(artist = word(artists, 1, sep = ",")) |>
                          select(-artists) 


                    
global_spotify_2023    
```

## Import & Wrangle Dataset 3 - World Data 2023 (Kaggle)

```{r}

## Load the world_data_2023
world_data_2023 <- read.csv('/Users/Jaykay/Desktop/INF 385T.9 - Data Wrangling/Final Project Data/world_data_2023.csv')

world_data_2023
```

```{r}

## Select County and Abbreviation columns only
## and assign it to a new table
abbreviation_table <- world_data_2023 |>
                        select(Country, Abbreviation)

abbreviation_table
```

## Import & Wrangle Dataset 4 - Weekly Charts of October 16, 2025 (Kworb)

References:

<https://stringr.tidyverse.org/reference/case.html>

<https://stringr.tidyverse.org/reference/word.html>

<https://stringr.tidyverse.org/reference/str_trim.html>

### New Zealand Dataset

```{r}

## Import New Zealand's weekly chart of 10/16/2025
`10_16_25_new_zealand` <- read.csv('/Users/Jaykay/Desktop/INF 385T.9 - Data Wrangling/Final Project Data/10_16_25_new_zealand.csv')

`10_16_25_new_zealand`
```

```{r}

## Remove all columns except for the Artist.and.Title column.
## Separate artist and song title from the Artist.and.Title column
new_zealand_25 <- `10_16_25_new_zealand` |>
                      select(Artist.and.Title) |>
                      separate_wider_delim(Artist.and.Title, 
                                           delim = "-",
                                           names = c("Artist", "Title"),
                                           too_many = "merge") 
```

```{r}

## Only keep the first artist
## For smoother joins, change following: 
## Remove any white space before and after the title and artist columns
## Change the case to all lowercase for both title and artist columns
new_zealand_2025 <- new_zealand_25 |>
                      mutate(Artist = word(Artist, 1, sep = ",")) |>
                      mutate(Artist = str_to_lower(str_trim(Artist)),
                             Title = str_to_lower(str_trim(Title)))

new_zealand_2025
```

### South Korea Dataset

```{r}

## Import South Korea's weekly chart of 10/16/2025
`10_16_25_korea` <-read.csv('/Users/Jaykay/Desktop/INF 385T.9 - Data Wrangling/Final Project Data/10_16_25_korea.csv')

`10_16_25_korea`
```

```{r}

## Remove all columns except for the Artist.and.Title 
## Separate artist and song title from the Artist.and.Title column
korea_25 <- `10_16_25_korea` |>
                      select(Artist.and.Title) |>
                      separate_wider_delim(Artist.and.Title, 
                                           delim = "-",
                                           names = c("Artist", "Title"),
                                           too_many = "merge")

korea_25
```

```{r}

## Only keep the first artist
## For smoother joins, change following: 
## Remove any white space before and after the title and artist columns
## Change the case to all lowercase for both title and artist columns
south_korea_2025 <- korea_25 |>                      
                      mutate(Artist = word(Artist, 1, sep = ",")) |>
                      mutate(Artist = str_to_lower(str_trim(Artist)),
                             Title = str_to_lower(str_trim(Title)))  

south_korea_2025
```

### USA Dataset

```{r}

## Import USA's weekly chart of 10/16/2025
`10_16_25_usa` <- read.csv('/Users/Jaykay/Desktop/INF 385T.9 - Data Wrangling/Final Project Data/10_16_25_usa.csv')

`10_16_25_korea`
```

```{r}

## Import weekly chart of USA and remove all columns 
## except for the Artist.and.Title column
## Separate artist and song title from the Artist.and.Title column
## Only keep the first artist
## Remove any white space before and after string
## Change both columns to all lowercase for smoother joining process.
us_25 <- `10_16_25_usa` |>
               select(Artist.and.Title) |>
               separate_wider_delim(Artist.and.Title, 
                                    delim = "-",
                                    names = c("Artist", "Title"),
                                    too_many = "merge")

us_25
```

```{r}

## Only keep the first artist
## For smoother joins, change following: 
## Remove any white space before and after the title and artist columns
## Change the case to all lowercase for both title and artist columns
usa_2025 <- us_25 |>
              mutate(Artist = word(Artist, 1, sep = ",")) |>
              mutate(Artist = str_to_lower(str_trim(Artist)),
                     Title = str_to_lower(str_trim(Title))) 

usa_2025
```

## Import & Wrangle Dataset 5 - Universal Top Spotify Songs (Kaggle)

This dataset was last updated on June 11, 2025.

References:

<https://readr.tidyverse.org/reference/type_convert.html>

```{r}

## Import the Universal Top Spotify Songs dataset
universal_top_spotify_songs <- read.csv('/Users/Jaykay/Desktop/INF 385T.9 - Data Wrangling/Final Project Data/universal_top_spotify_songs.csv')

universal_top_spotify_songs
```

```{r}

## Import the Universal Top Spotify Songs dataset
## Keep following columns only:
## name, artists, danceability, energy, loudness, liveness, tempo
universal_spotify <- universal_top_spotify_songs |>
                      select(name, artists, danceability, energy, loudness, valence, tempo)|>
                      rename(title = name)
    
universal_spotify
```

```{r}

## Split the artists column and only keep the first artist
## Change both columns to all lowercase for smoother joining process.
spotify_2025_R <- universal_spotify |>
                    mutate(artists = str_to_lower(str_trim(as.character(artists))), 
                             title = str_to_lower(str_trim(title))) |>
                    mutate(artist = word(artists, 1, sep = ",")) |>
                    select(-artists) |>
                    relocate(artist, .after = title)

spotify_2025_R
```

## Join dataset 2 and 3 - replace country abbreviation with its full name

```{r}

## Match spotify's country abbreviation with abbreviation table's abbreviation
## Drop abbreviation column
## Rearrange country column to the front
## Remove any unmatched rows
spotify_global_2023_matched <- global_spotify_2023 |>
                                left_join(abbreviation_table,
                                          join_by(country == Abbreviation)) |>
                                select(!country) |>
                                drop_na() |>
                                relocate(Country, everything())
                                

spotify_global_2023_matched
```

## Calculate mean danceability, energy, loudness, valence, and tempo by Country for 2023 Spotify Dataset

```{r}

## Calculate averages of spotify_global_2023_matched 
## dataset by Country
spotify_global_2023_avg <- spotify_global_2023_matched |>
                            group_by(Country) |>
                            summarise(danceability_avg = mean(danceability),
                                      energy_avg = mean(energy),
                                      loudness_avg = mean(loudness),
                                      valence_avg = mean(valence),
                                      tempo_avg = mean(tempo))

spotify_global_2023_avg
```

## Join 2025 Spotify Dataset to Country Datasets for Analysis

References:

<https://dplyr.tidyverse.org/reference/distinct.html>

### New Zealand Dataset

```{r}

## Join New Zealand dataframe and updated Spotify dataframe
## Left join to keep all the rows of left dataframe
new_zealand_2025_matched_r <- new_zealand_2025 |>
                                left_join(spotify_2025_R,
                                          join_by(Artist == artist,
                                                  Title == title))
new_zealand_2025_matched_r
```

```{r}

## Drop duplicate rows and songs with no analytic data
new_zealand_2025_matched_r_cleaned <- new_zealand_2025_matched_r |>
                                        distinct(Artist, Title, .keep_all = TRUE) |>
                                        drop_na(danceability)

new_zealand_2025_matched_r_cleaned
```

```{r}

## Calculate average for each column
## Add Country and year columns for identification
## Will be used later for 2023 and 2025 comparison
new_zealand_spotify_2025_avg <- new_zealand_2025_matched_r_cleaned |>
                                  summarise(danceability_avg = mean(danceability),
                                            energy_avg = mean(energy),
                                            loudness_avg = mean(loudness),
                                            valence_avg = mean(valence),
                                            tempo_avg = mean(tempo)) |>
                                  mutate(Country = "New Zealand",
                                         year = 2025) |>
                                  relocate(Country, year, everything())

new_zealand_spotify_2025_avg
```

### South Korea Dataset

```{r}

## Join South Korea dataframe and updated Spotify dataframe
## Left join to keep all the rows of left dataframe
south_korea_2025_matched_r <- south_korea_2025 |>
                                left_join(spotify_2025_R,
                                          join_by(Artist == artist,
                                                  Title == title))
south_korea_2025_matched_r
```

```{r}

## Remove duplicate 
## Drop title and artist columns from spotify_2025_py_cleaned
## Drop duplicate rows and songs with no analytic data
south_korea_2025_matched_r_cleaned <- south_korea_2025_matched_r |>
                                        distinct(Artist, Title, .keep_all = TRUE) |>
                                        drop_na(danceability)

south_korea_2025_matched_r_cleaned
```

```{r}

## Calculate average for each column
## Add Country and year columns for identification
## Will be used later for 2023 and 2025 comparison
south_korea_spotify_2025_avg <- south_korea_2025_matched_r_cleaned |>
                                  summarise(danceability_avg = mean(danceability),
                                            energy_avg = mean(energy),
                                            loudness_avg = mean(loudness),
                                            valence_avg = mean(valence),
                                            tempo_avg = mean(tempo)) |>
                                  mutate(Country = "South Korea",
                                         year = 2025) |>
                                  relocate(Country, year, everything())

south_korea_spotify_2025_avg
```

### USA Dataset

```{r}

## Join USA dataframe and updated Spotify dataframe
## Left join to keep all the rows of left dataframe
usa_2025_matched_r <- usa_2025 |>
                        left_join(spotify_2025_R,
                                  join_by(Artist == artist,
                                          Title == title))

usa_2025_matched_r
```

```{r}

## Remove duplicate 
## Drop title and artist columns from spotify_2025_py_cleaned
## Drop duplicate rows and songs with no analytic data
usa_2025_matched_r_cleaned <- usa_2025_matched_r |>
                                        distinct(Artist, Title, .keep_all = TRUE) |>
                                        drop_na(danceability)

usa_2025_matched_r_cleaned
```

```{r}

## Calculate average for each column
## Add Country and year columns for identification
## Will be used later for 2023 and 2025 comparison
usa_spotify_2025_avg <- usa_2025_matched_r_cleaned |>
                          summarise(danceability_avg = mean(danceability),
                                    energy_avg = mean(energy),
                                    loudness_avg = mean(loudness),
                                    valence_avg = mean(valence),
                                    tempo_avg = mean(tempo)) |>
                          mutate(Country = "United States",
                                 year = 2025) |>
                          relocate(Country, year, everything())

usa_spotify_2025_avg
```

## Correlation Studies

References:

<https://www.geeksforgeeks.org/r-language/how-to-plot-a-correlation-matrix-into-a-graph-using-r/>

<https://www.geeksforgeeks.org/r-language/correlation-matrix-in-r-programming/>

<https://r-charts.com/correlation/heat-map-ggplot2/>

<https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>

<https://collinn.github.io/teaching/2023/labs/lab4.htm>

<https://stackoverflow.com/questions/73723000/how-to-add-a-caption-subtitle-to-corrplot-plot-r>

<https://www.rdocumentation.org/packages/corrplot/versions/0.95/topics/corrplot>

```{r}

## Import libraries needed for correlation analysis
library(ggplot2)
library(corrplot)
library(ggcorrplot)
library(PerformanceAnalytics)
library(reshape2)
```

### Preparing Datasets for Correlation Analysis

```{r}

## Combine aggregated Spotify dataset with happiness report
## Drop any countrys with no happiness score
happiness_spotify_2023 <- spotify_global_2023_avg |>
                                left_join(happiness_report_23, 
                                join_by(Country == country_name))|>
                                drop_na()

happiness_spotify_2023
```

```{r}
## Remove Country column for correlation analysis
happiness_spotify_2023_cleaned <- happiness_spotify_2023 |>
                                        select(!Country)

happiness_spotify_2023_cleaned
```

```{r}

## Calculate Correlation Matrix
cor_matrix_happiness_spotify_23 <- cor(happiness_spotify_2023_cleaned, use = "complete.obs")

## Melt the matrix for readability
melted_matrix <- melt(cor_matrix_happiness_spotify_23)

melted_matrix
```

### Analyzing Correlations in the Top 10 Happiness Score Countries

```{r}

## Left_join top_happiness_score_23 with spotify_global_2023_avg
top_happiness_score_spotify_23 <- happiness_spotify_2023 |>
                                    arrange(desc(happiness_score)) |>
                                    slice_head(n = 10)
                                    

top_happiness_score_spotify_23
```

```{r}

## Remove Country column for correlation study
top_happiness_countries_23 <- top_happiness_score_spotify_23 |>
                                select(!Country)


## Find correlation for Finland data
## Melt the correlation table for ggplot
top_happiness_countries_23_corr <- cor(top_happiness_countries_23)

top_happiness_countries_23_corr_melt <- melt(top_happiness_countries_23_corr)

top_happiness_countries_23_corr_melt
```

```{r}

ggplot(data = top_happiness_countries_23_corr_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", 
                       high = "blue", 
                       mid = "white", 
                       midpoint = 0, 
                       limit = c(-1,1), 
                       space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1, 
                                   size = 9)) +
  labs(title = "Correlations for the Top 10 Happiest Countries") +
  coord_fixed()
```

```{r}

## Visualize correlation matrix as a color chart with correlation values
corrplot(top_happiness_countries_23_corr,
         method = "number",
         title = "Correlations for the Top 10 Happiness Score Countries",
         mar = c(1,5,4,2),
         tl.srt = 45,
         tl.col = "black",
         tl.cex = 0.6,
         number.cex = 0.5,
         col = COL2('RdBu', n = 200))
```

### Analyzing Correlations in the Bottom 10 Happiness Score Countries

```{r}

## Left_join top_happiness_score_23 with spotify_global_2023_avg
bottom_happiness_score_spotify_23 <- happiness_spotify_2023 |>
                                      arrange(happiness_score) |>
                                      slice_head(n = 10)

bottom_happiness_score_spotify_23
```

```{r}

## Remove Country column for correlation study
bot_happiness_countries_23 <- bottom_happiness_score_spotify_23 |>
                                select(!Country)


## Find correlation for Finland data
## Melt the correlation table for ggplot
bot_happiness_countries_23_corr <- cor(bot_happiness_countries_23)

bot_happiness_countries_23_corr_melt <- melt(bot_happiness_countries_23_corr)

bot_happiness_countries_23_corr_melt
```

```{r}

## Visualize correlation  as a heatmap 
ggplot(data = bot_happiness_countries_23_corr_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", 
                       high = "blue", 
                       mid = "white", 
                       midpoint = 0, 
                       limit = c(-1,1), 
                       space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1, 
                                   size = 9)) +
  labs(title = "Correlations for the Bottom 10 Happiness Score Countries") +
  coord_fixed()
```

```{r}

## Visualize correlation as a correlation values chart 
corrplot(bot_happiness_countries_23_corr,
         method = "number",
         title = "Correlations for the Bottom 10 Happiness Score Countries",
         mar = c(1,5,4,2),
         tl.srt = 45,
         tl.col = "black",
         tl.cex = 0.6,
         number.cex = 0.5,
         col = COL2('RdBu', n = 200))
```

## Comparing 2023 and 2025 Spotify Dataset

Since 2025 Happiness Report is not available yet, we will compare the average of music analytic components and note if theres any changes.

Tools/Reference:

<https://dplyr.tidyverse.org/reference/bind_rows.html>

<https://stackoverflow.com/questions/30219908/change-row-text-using-dplyr?utm_source=chatgpt.com>

<https://www.r-bloggers.com/2016/09/reshape-r-package-reshape2-melt-and-cast/?utm_source=chatgpt.com>

### New Zealand

```{r}

## Filter 2023 New Zealand spotify data
## Mutate Country column so that it is now New Zealand 2023
new_zealand_2023 <- filter(spotify_global_2023_avg,
                           Country == "New Zealand") |>
                    mutate(year = 2023)
                            
new_zealand_2023
```

```{r}

## Concat 2023 and 2025 New Zealand data into one table
## Melt the table along the Country and year
new_zealand_spotify_combined <- bind_rows(new_zealand_2023, new_zealand_spotify_2025_avg) |>
                                  relocate(Country, year, everything())

new_zealand_spotify_combined_melt <- melt(new_zealand_spotify_combined,
                                          id.vars = c("Country", "year"),
                                          measured.vars = c("danceability_avg",
                                                            "energy_avg",
                                                            "loudness_avg",
                                                            "liveness_avg",
                                                            "valence_avg",
                                                            "tempo_avg"))

new_zealand_spotify_combined_melt
```

### South Korea

```{r}

## Filter 2023 New Zealand spotify data
## Mutate Country column so that it is now New Zealand 2023
south_korea_2023 <- filter(spotify_global_2023_avg,
                           Country == "South Korea") |>
                    mutate(year = 2023)
                            
south_korea_2023
```

```{r}

## Concat 2023 and 2025 New Zealand data into one table
## Melt the table along the Country and year
south_korea_spotify_combined <- bind_rows(south_korea_2023, south_korea_spotify_2025_avg) |>
                                  relocate(Country, year, everything())

south_korea_spotify_combined_melt <- melt(south_korea_spotify_combined,
                                          id.vars = c("Country", "year"),
                                          measured.vars = c("danceability_avg",
                                                            "energy_avg",
                                                            "loudness_avg",
                                                            "liveness_avg",
                                                            "valence_avg",
                                                            "tempo_avg"))

south_korea_spotify_combined_melt
```

### USA

```{r}

## Filter 2023 New Zealand spotify data
## Mutate Country column so that it is now New Zealand 2023
usa_2023 <- filter(spotify_global_2023_avg,
                   Country == "United States") |>
            mutate(year = 2023)
                            
usa_2023
```

```{r}

## Concat 2023 and 2025 New Zealand data into one table
## Melt the table along the Country and year
usa_spotify_combined <- bind_rows(usa_2023, usa_spotify_2025_avg) |>
                                  relocate(Country, year, everything())

usa_spotify_combined_melt <- melt(usa_spotify_combined,
                                          id.vars = c("Country", "year"),
                                          measured.vars = c("danceability_avg",
                                                            "energy_avg",
                                                            "loudness_avg",
                                                            "liveness_avg",
                                                            "valence_avg",
                                                            "tempo_avg"))

usa_spotify_combined_melt
```
